---
title: "Data2015"
author: "B"
date: "Tuesday, June 02, 2015"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, cache=FALSE} 
#########function to install (if needed) and load R packages by list
libs<-c("readxl","dplyr") #list of packages to load

installLoad<-function(pck){ #user defined function
    if(!pck%in%installed.packages()){install.packages(pck)}
    require(pck, character.only = TRUE)
  }
lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

To Do List
-------------------------
* Check POSIXct values for Dates and Times


Introduction
-------------------------
During the summer of 2015 the New England states monitored in lakes.  Participants included state and tribal governments, local watershed associations, EPA Region 1, and the EPA Atlantic Ecology Division. 


Data Definitions
-------------------------

Field  | Units | Description
------------- | ------------- | -------------
**Order:**|(integer)|Unique Identifier for sample by Organization-disregard this is for house keeping only.
**YourName:**|(text)|Enter the name of the person entering the data
**Organization:**|(text)|Enter the name of the organization responsible for collecting the samples-e.g. Vermont DEC
**SiteID:**|(text)|If the location has a site ID from your organization enter it here
**WaterbodyName:**|(text)|Enter the name of the lake: be consistent with spelling and capitalization
**State:**|(lookup)|Choose the two letter state code from the drop down list
**SiteLocation:**|(text)|Some organizations have site names within a lake.  Enter these here.
**SampleYear:**|(YYYY)|Year sample was collected in format YYYY (e.g., 2014)
**SampleMonth:**|(MM)|Month sample was collected in integer format (e.g., months 1 to 12 )
**SampleDay:**|(DD)|Day sample was collected in integer format (e.g., day 1 to 31 )
**NameOfSamplers:**|(text)|Add the names of the field crew separated by commas.
**WeatherConditions:**|(text)|General Weather conditions separated by commas.  E.g., Cloudy, Windy, Cold
**SampleLocation:**|(lookup)|Where (in the lake) was the sample collected followed by the replicate number? WithinLake=WL1, WL2, or WL3; ShoreSide=SS1, SS2, SS3; Can also add Calibration or Blank for validation readings or Other.
**SampleMethod:**|(lookup)|How was the sample collected? Grab = Grab sample for surface blooms; Composite =Composite; Integrated = Integrated tube sample; Validation = Use this for Blanks and Calibration Standards; Other = give details in the comments section.
**Depth:**|(integer)|Within Lake samples are at 3m and Shoreside are at 1m.  If samples taken at different depths not the depth here and enter details in the comments section.  Leave blank for standards and blanks.
**Latitude:**|(decimal)|Latitude in Decimal Degrees with 4 decimal places
**Longitude:**|(decimal)|Longitude in Decimal Degrees with 4 decimal places
**SampleHour:**|(HH)|Hour sample was taken in 24 hour format
**SampleMinutes:**|(MM)|Minute sample was taken in integer format
**Filtered:**|(TRUE/FALSE)|Was the sample filtered?
**Frozen:**|(TRUE/FALSE)|Was the sample frozen?
**Parameter:**|(lookup)|Phycocyanin or Chlorophyll?
**Value:**|(decimal)|Reading from the fluorometer
**Units:**|(lookup)|What were the units recorded.  If not "RFU" or "µ/L" flag the entry and add comment with the units used. 
**Rep:**|(integer)|If you made more than one measurement per sample or took more than one sample per site assign a replicate number to each observations and add notes in the comment field.
**Fluorometer:**|(lookup)|Type of fluorometer used?
**AnalysisYear:**|(YYYY)|Year sample was analyzed in format YYYY (e.g., 2014)
**AnalysisMonth:**|(MM)|Month sample was analyzed in integer format (e.g., months 1 to 12 )
**AnalysisDay:**|(DD)|Day sample was analyzed in integer format (e.g., day 1 to 31 )
**AnalysisHour:**|(HH)|Hour sample was analyzed in 24 hour format
**AnalysisMinutes:**|(MM)|Minute sample was analyzed in integer format
**GPSType:**|(text)|How was the location determined.  GPS (type), map, or google?
**Photos:**|(TRUE/FALSE)|Where photos taken?
**Flag:**|(TRUE/FALSE)|Add a flag for any data line that needs further validation or processing
**Comments:**|(text)|Add details of flags or any notes about the data line or sample.
**ID:**|(integer)|Unique Identifier for Sample
**LocID:**|(integer)|Unique Identifier for the Location based on unique Lon/Lat combinations

Data Sources
-------------------------
Information on how the data were provided by the States and Organizations that participated in the monitoring.


### CRWA

* Monitoring data received from Elisabeth Cianciola on 10/16/2015 (CRWA_2015_EPA_CyanoData.xls)
* Secondary standards log received from Elisabeth Cianciola on 11/5/2015 (CRWA_SecondaryStandardsLog2015.xlsx)

**Questions for Elisabeth sent 12/2/15
* It looks like one of your stationIDs was entered differently for the lab and the field data.  Can you confirm that stationID="BROAD" for the field data is the same as stationID="BR" for the lab data?
* It appears that the field data for samples taken during August, September, and October were not entered into the spreadsheet.  I have added the missing observations (highlighted) to the attached spreadsheet.  Could you check these and add the missing information?  


**Data steps**

* read the data 
* One stationID is differenct for field and lab data.  For the field data stationID='BROAD' is listed as stationID=BR in the lab data. So the field data values for stationID were changed to match the lab data.
*No field data for samples taken 8/20, 8/31, 9/10, 9/17, 9/24, 10/01, & 10/08

```{r CRWA, include=FALSE, echo=FALSE, cache=FALSE} 
#read field data 
  Field<-read_excel('Data2015/rawData/CRWA_2015_EPA_CyanoData.xls',sheet='FieldDataEntry')

#read Lab data 
  Lab<-read_excel('Data2015/rawData/CRWA_2015_EPA_CyanoData.xls',sheet='LabDataEntry')

#change Field$stationID=='BROAD' to Field$stationID=='BR'
  Field$stationID[Field$stationID=='BROAD']<-'BR'

#merge field and lab data
  CRWA<-full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate'))

#check to make sure the join was successful
  #print a few columns
    select(CRWA,waterbodyID,stationID,sampleID,sampleDate,analysisID,analysisDate)

  #print rows where the are no lab data to match the field data
    #CRWA[is.na(CRWA$analysisID),c('orgID','waterbodyID','stationID','sampleID','sampleDate','analysisID','analysisDate')]
    filter(CRWA,is.na(analysisID)) %>%
      select(waterbodyID,stationID,sampleID,sampleDate,analysisID,analysisDate)
           
          
    ####all good

  #print rows where the are no field data to match the lab data
    filter(CRWA,is.na(orgID)) %>%
      select(orgID,waterbodyID,stationID,sampleID,sampleDate,analysisID,analysisDate)
    
    #No field data for samples taken 8/20, 8/31, 9/10, 9/17, 9/24, 10/01, & 10/08
      filter(CRWA,is.na(orgID)) %>%
        select(orgID,sampleDate) %>%
          distinct()
```

### MADEP

* Monitoring data received from Joan Beskenis on 11/16/2015 (cyanoMon2015DataEntrynewestjb.xlsx)
* Secondary standards log received from Joan Beskenis on 11/10/2015 (secondaryStandardsLog2015 jb(2).xlsx)
    - Note original sheet did not include the "Production lot number" & "set number".  She sent this in a later email and I added it to the spreadsheet. Production lot #="Rhod-1001";"Set #="1001-D""


**Data steps**

* read the data 
* visual inspection of data shows that 3 samples (analysisID == 4, 5, & 6) have fluorometerType=='Beagle' but commentSample=='no Beagle, cyclops probe'; fluorometerType changed to 'Other'
* unidiluted samples listed as dilution=='n';changed to dilution==NA
* no lab data for Jamaica Pond and White Pond.  This is noted in the comments


```{r MADEP, include=FALSE, echo=FALSE, cache=FALSE} 
#read field data 
  Field<-read_excel('Data2015/rawData/cyanoMon2015DataEntrynewestjb.xlsx',sheet='Field Data Entry') #15
#there are four lines with no data-remove these
  Field<-Field[!is.na(Field$waterbodyID),] #11

#read Lab data 
  Lab<-read_excel('Data2015/rawData/cyanoMon2015DataEntrynewestjb.xlsx',sheet='Lab Data Entry') #65541
#nrow(Lab)==65541; only 28 lines have data.  Keep these
  Lab<-Lab[!is.na(Lab$waterbodyID),] #28

#merge field and lab data
  MADEP<-full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate'))

#fluorometerType changes (see above)
MADEP<-mutate(MADEP,fluorometerType=replace(fluorometerType, analysisID%in%c(4:6), 'Other'))

#dilution changes (see above)
MADEP<-mutate(MADEP,dilution=replace(dilution, dilution=='n', 'No'))

#check to make sure the join was successful
  #print a few columns
    select(MADEP,waterbodyID,stationID,sampleID,sampleDate,analysisID,analysisDate)

  #print rows where the are no lab data to match the field data (see above)
    filter(MADEP,is.na(analysisID)) %>%
      select(waterbodyID,stationID,sampleID,sampleDate,analysisID,analysisDate)
    View(filter(MADEP,is.na(analysisID)))
           
  #print rows where the are no field data to match the lab data (none)
    filter(MADEP,is.na(orgID)) %>%
      select(orgID,waterbodyID,stationID,sampleID,sampleDate,analysisID,analysisDate)
    #View(filter(MADEP,is.na(orgID)))
    
```

### NHDES

* Monitoring data received from Sonya Carlson on 10/08/2015 (NH-for-EPA-2015-cyanoMonDataEntry.xlsx)
* Some additional monitoring data received from Sonya on 10/16/2015 (NH-for-EPA-2015-cyanoMonDataEntry-addenda.xlsx)
* A few more lines of monitoring data received from Sonya on 10/29/2015 (NH-for-EPA-2015-cyanoMonDataEntry-addenda2.xlsx)
* Secondary standards log received from Sonya on 11/13/2015 (NH-secondaryStandardsLog2015.xlsx)

**Check**
* missing lab data for six samples
* missing field data for 76 lab readings

**Data steps**

* read and combine the 3 sets of monitoring data
* remove NA rows
* The field data for 10/16/2015 has only 1 observation and the sampleTimeis missing.  To merge this with the other field records need to convert the sampleTime column to POSIXct



```{r NHDES, include=FALSE, echo=FALSE, cache=FALSE} 
#read field data 
  Field<-read_excel('Data2015/rawData/NH-for-EPA-2015-cyanoMonDataEntry.xlsx',sheet='Field Data Entry') 
  table(is.na(Field$waterbodyID)) #only 270 lines have data; keep these
  Field<-Field[!is.na(Field$waterbodyID),];nrow(Field) #270
#get second set of field data
  Field2<-read_excel('Data2015/rawData/NH-for-EPA-2015-cyanoMonDataEntry-addenda.xlsx',sheet='Field Data Entry') 
  table(is.na(Field2$waterbodyID)) #only 1 line have data; keep it
  Field2<-Field2[!is.na(Field2$waterbodyID),];nrow(Field2) #1
#the sampleTime for the one Field2 observation is missing.  Need to convert this to POSIXct
  Field2$sampleTime<-as.POSIXct(Field2$sampleTime, origin = "1960-01-01") 
#get third set of field data
  Field3<-read_excel('Data2015/rawData/NH-for-EPA-2015-cyanoMonDataEntry-addenda2.xlsx',sheet='Field Data Entry') 
  table(is.na(Field3$waterbodyID)) #28 lines have data; keep em
  Field3<-Field3[!is.na(Field3$waterbodyID),];nrow(Field3) #28
#combine field,field2 & field3
  Field<-rbind(Field, Field2,Field3);nrow(Field)

#read Lab data 
  Lab<-read_excel('Data2015/rawData/NH-for-EPA-2015-cyanoMonDataEntry.xlsx',sheet='Lab Data Entry') 
  names(Lab) #the last two columns were added by Sonya-delete these
  Lab<-Lab[,1:17]
  table(is.na(Lab$waterbodyID)) #14 lines have are blank; delete these
  Lab<-Lab[!is.na(Lab$waterbodyID),];nrow(Lab) #2115
#read 2nd set of lab data
  Lab2<-read_excel('Data2015/rawData/NH-for-EPA-2015-cyanoMonDataEntry-addenda.xlsx',sheet='Lab Data Entry')
  table(is.na(Lab2$waterbodyID)) #all 9 lines have data
#read 3rd set of lab data
  Lab3<-read_excel('Data2015/rawData/NH-for-EPA-2015-cyanoMonDataEntry-addenda2.xlsx',sheet='Lab Data Entry')
  table(is.na(Lab3$waterbodyID)) #all 177 lines have data
  names(Lab3) #the last two columns were added by Sonya-delete these
  Lab3<-Lab3[,1:17]
#combine Lab,Lab2 & Lab3
  Lab<-rbind(Lab, Lab2,Lab3);nrow(Lab)#2301 (2115+177+9)

#merge field and lab data
  NHDES<-full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate'))

#check to make sure the join was successful
  #view rows where the are no lab data to match the field data (see above)
    View(filter(NHDES,is.na(analysisID)))
           
  #view rows where the are no field data to match the lab data (none)
    View(filter(NHDES,is.na(orgID)))    
```


### LIMTEX

* Monitoring data received from Nancy Leland on 11/15/2015 (Copy of cyanoMon2015DataEntry.xls)
* Secondary standards log received from Nancy Leland on 11/15/2015 (Copy of secondaryStandardsLog2015.xlsx)


**Data steps**

* read the data 
* complete match between lab and field data


```{r LIMTEX, include=FALSE, echo=FALSE, cache=FALSE} 
#read field data 
  Field<-read_excel('Data2015/rawData/Copy of cyanoMon2015DataEntry.xls',sheet='Field Data Entry') 
  table(is.na(Field$waterbodyID)) #there are 3 lines with no data-remove these
  Field<-Field[!is.na(Field$waterbodyID),];nrow(Field) #10

#read Lab data 
  Lab<-read_excel('Data2015/rawData/Copy of cyanoMon2015DataEntry.xls',sheet='Lab Data Entry') 
  table(is.na(Lab$waterbodyID)) #20 lines have data; keep these
  Lab<-Lab[!is.na(Lab$waterbodyID),];nrow(Lab) #20

#merge field and lab data
  LIMTEX<-full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate'))

#check to make sure the join was successful
  #view rows where the are no lab data to match the field data (see above)
    View(filter(LIMTEX,is.na(analysisID))) #all data match
           
  #view rows where the are no field data to match the lab data (none)
    View(filter(LIMTEX,is.na(orgID)))  #all data match  
    
```

### CTDEEP

* Monitoring data received from Tracy Lizotte on 12/10/2015 (cyanoMon2015DataEntry- CT DEEP.xls)
* A key to photos was received from Tracy Lizotte on 12/10/2015 (Cyanobacteria ID Legend 2015.xlsx)
    - NOTE: photos have not arrived yet but will
* Secondary standards log NOT received 

**Check**
* missing lab data for 68 samples
* missing field data for 274 lab readings


**Data steps**

* read the data 
* waterbodyID and stationID are the same so I guess there is one station per lake
* waterbodyID and stationID read as character for the field data and numeric for the Lab data
* For the 8 occurences of waterbodyID==15916-M read as NA
* Change Lab$waterbodyID & Lab$sampleID to as.character and replace the missing IDs


* Lab data read in with waterbodyID as numeric.  This converted 
* complete match between lab and field data


```{r CTDEEP, include=FALSE, echo=FALSE, cache=FALSE} 
#read field data 
  Field<-read_excel('Data2015/rawData/cyanoMon2015DataEntry- CT DEEP.xls',sheet='Field Data Entry') 
  View(Field)
  table(is.na(Field$waterbodyID)) #there are 70 lines of data

#read Lab data 
  Lab<-read_excel('Data2015/rawData/cyanoMon2015DataEntry- CT DEEP.xls',sheet='Lab Data Entry')
  View(Lab)
#delete rows with missing data
  Lab<-Lab[Lab$analysisID<284,]
#convert waterbodyID and stationID to character & replace missing values
  Lab$waterbodyID<-as.character(Lab$waterbodyID)
  Lab$waterbodyID[is.na(Lab$waterbodyID)]<-'15916-M'
  Lab$stationID<-as.character(Lab$stationID)
  Lab$stationID[is.na(Lab$stationID)]<-'15916-M'

#merge field and lab data
  CTDEEP<-full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate'))
    

#check to make sure the join was successful
  #view rows where the are no lab data to match the field data (see above)
    View(filter(CTDEEP,is.na(analysisID))) #all data match
           
  #view rows where the are no field data to match the lab data (none)
    View(filter(CTDEEP,is.na(orgID)))  #all data match  
    
```

