---
title: "Data2015"
author: "B"
date: "Tuesday, June 02, 2015"
output: html_document
---
<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  library(knitr)
    setwd("Data2015")
    knit('Data2015.rmd')
-->

```{r setup, include=FALSE, echo=FALSE, cache=FALSE} 
#########function to install (if needed) and load R packages by list
libs<-c("xlsx","dplyr") #list of packages to load

installLoad<-function(pck){ #user defined function
    if(!pck%in%installed.packages()){install.packages(pck)}
    require(pck, character.only = TRUE)
  }
lapply(libs,function(x) installLoad(x))  #Load/Install require packages

#load functions
source('../functions/readField2015.r')
source('../functions/readLab2015.r')
```

To Do List
-------------------------
* Check POSIXct values for Dates and Times
* Convert yes/no to TRUE/FALSE
* edit Data Definitions to 2015 values
* CRWA corrections
* NHDES corrections


Introduction
-------------------------
During the summer of 2015 the New England states monitored in lakes.  Participants included state and tribal governments, local watershed associations, EPA Region 1, and the EPA Atlantic Ecology Division. 


Data Definitions
-------------------------

Field  | Description
------------- | -------------
**orgID:**|Acronym for the Organization collecting data (e.g., AED, CRWA, CTDEEP, MEDEP, NHDES, RIWW, UNH, VTDEC)
**ContactName:**|Name of contact person for data entry and questions.
**Email:**|Email address of contact person
**Phone:**|Phone Number of contact person
**waterbodyID:**|Important: if your organzation has a unique identifier for the waterbody use it here.  If not use your orgID followed by an underscore and a uniuqe three digit  number. (e.g. UNH_001)
**waterbodyName:**|Waterbody Name with no abbreviations.  Check spelling and try to be consistent.
**state:**|Two letter state code
**town:**|Closest town to the waterbody (if known)
**commentWB:**|Any comments or observations about the waterbody go here in text format.
**stationID:**|Important: if your organzation has a unique identifier for the station use it here.  If not use "Sta" followed by a unique two digit  number.  (e.g. Sta01, Sta02, etc).
**stationDescription:**|Text description of the station location; something to help you remember where it is.
**stationType:**|Select "nearShore" or "offShore"
**longitudeSta:**|Longitude in decimal degrees (WGS84) of the station. Miniumum of 4 decimal places; 6 decimal places prefered.
**latitudeSta:**|Latitude in decimal degrees (WGS84) of the station. Miniumum of 4 decimal places; 6 decimal places prefered.
**locationSourceSta:**|How was the location determined (e.g.,"WaterbodyDatabase"; "GPS"; "GoogleEarth"; "BingMaps"; "topoMap")?
**commentStation:**|Any comments or observations about the station go here in text format.
**sampleID:**|Assign a unique ID to the sample.  Could be  "Samp" and a two digit sequential number (e.g., Samp01, Samp02 etc.)
**sampleDate:**|Date the sample was taken in format MM/DD/YYYY
**sampleTime:**|Time the sample was taken in format HH:MM AM/PM
**fieldCrew:**|Names of the field crew separated by commas (e.g., Hilary Snook, Jim Haney)
**sampleMethod:**|This should only be "Integrated Sampler"  but if you do use another method let us know.
**sampleRep:**|Select "primary" or "duplicate".  For each station and date you can have only one primary sample but many duplicates.
**sampleDepthM:**|Sample depth in Meters.  This should be 1 (m) for "nearShore" stations and 3 (m) but other depths could be entered if necessary.
**waterTempC:**|Enter the water temperature (Centigrade) at sample site 
**photoSample:**|Did you take photos during the sampling? Choose yes or no
**surfaceWaterCondition:**|Choices restricted to  "Calm"; "Ripples"; "Choppy"; &  "White Caps"
**weather:**|Choices restricted to  "Clear"; "Partly Cloudy"; "Overcast"; & "Rain"
**commentSample:**|Any comments or observations about the sample go here in text format.
**analysisID:**|If you have your own analysis ID use it here otherwise sequentially number the analyses
**analysisDate:**|Date the sample was analyzed in format MM/DD/YYYY
**analystName:**|Name of the person in charge of the analysis
**frozen:**|Was the sample frozen before analysis?
**filtered:**|Was the sample filtered before analysis?
**dilution:**|Dilution factor for samples.  Use 1:1 for undiluted; 1:10 for a 1 to 10 dilution; etc.
**sampleTemperatureC:**|Enter the temperature (Centigrade) of the sample just prior to taking the fluorometry reading
**ChlaUGL:**|Fluorometer reading in ug/l for Chlorophyll
**PhycoUGL:**|Fluorometer reading in ug/l for Phycocyanin
**analysisRep:**|Select "primary" or "duplicate".  For each sample you can have only one primary reading but many duplicates.
**fluorometerType:**|This should be "Beagle".
**photosAnalysis:**|Did you take photos during the analysis? Could include microscope photos? Choose yes or no
**commentAnalysis:**|Any comments or observations about the analysis go here in text format.


Data Sources
-------------------------
Information on how the data were provided by the States and Organizations that participated in the monitoring.


### CRWA

* Monitoring data received from Elisabeth Cianciola on 10/16/2015 (CRWA_2015_EPA_CyanoData.xls)
* Secondary standards log received from Elisabeth Cianciola on 11/5/2015 (CRWA_SecondaryStandardsLog2015.xlsx)

**Questions for Elisabeth sent 12/2/15**
* It looks like one of your stationIDs was entered differently for the lab and the field data.  Can you confirm that stationID="BROAD" for the field data is the same as stationID="BR" for the lab data?
* It appears that the field data for samples taken during August, September, and October were not entered into the spreadsheet.  I have added the missing observations (highlighted) to the attached spreadsheet.  Could you check these and add the missing information?  

**Message to Elisabeth sent 01/07/16**
* Haven't heard back from Elisabeth so I sent new message that included a spreadsheet (CRWAdataToCheck.xls) for her to fill in.

**Data steps**

* read the data 
* One stationID is differenct for field and lab data.  For the field data stationID='BROAD' is listed as stationID=BR in the lab data. So the field data values for stationID were changed to match the lab data.
*No field data for samples taken 8/20, 8/31, 9/10, 9/17, 9/24, 10/01, & 10/08

```{r CRWA, include=FALSE, echo=FALSE, cache=FALSE} 
#set data input and output names
In<-'CRWA_2015_EPA_CyanoData.xls'
Out<-'CRWA'

#data log
dataReceived<-data.frame(Org=Out,monitoring=TRUE,standards=TRUE,Comments='Missing Field Data',stringsAsFactors=FALSE)

#read field data 
  Field<-readField2015(In)
  dim(Field) #18 28

#read Lab data 
  Lab<-readLab2015(In)
  dim(Lab) #37 17

#change Field$stationID=='BROAD' to Field$stationID=='BR'
  Field$stationID[Field$stationID=='BROAD']<-'BR'

#merge field and lab data
  assign(Out,full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate')))
  dim(get(Out)) #37, 41

#look for rows missing field or lab data 
    Chk<-filter(get(Out),is.na(orgID)|is.na(analysisID)) 
    if(nrow(Chk)==0){
      print("All records match")
        }else{
      select(Chk,waterbodyID, stationID,sampleID,sampleDate,analysisID)}#No field data for samples taken 8/20, 8/31, 9/10, 9/17, 9/24, 10/01, & 10/08

#output file to xls to send back for revisions
    File<-paste('rawData/',Out,'dataToCheck.xls',sep='')
    write.xlsx(Chk, file=File,row.names=FALSE,showNA=FALSE)
```

### MADEP

* Monitoring data received from Joan Beskenis on 11/16/2015 (cyanoMon2015DataEntrynewestjb.xlsx)
* Secondary standards log received from Joan Beskenis on 11/10/2015 (secondaryStandardsLog2015 jb(2).xlsx)
    - Note original sheet did not include the "Production lot number" & "set number".  She sent this in a later email and I added it to the spreadsheet. Production lot #="Rhod-1001";"Set #="1001-D""


**Data steps**

* read the data 
* visual inspection of data shows that 3 samples (analysisID == 4, 5, & 6) have fluorometerType=='Beagle' but commentSample=='no Beagle, cyclops probe'; fluorometerType changed to 'Other'
* no lab data for Jamaica Pond and White Pond.  This is noted in the comments
* Time format needs work.  Used 24hour format with AM/PM
* The times were manually corrected and the data resaved as: 'cyanoMon2015DataEntrynewestjb_edited.xlsx'


```{r MADEP, include=FALSE, echo=FALSE, cache=TRUE} 
#set data input and output names
In<-'cyanoMon2015DataEntrynewestjb_edited.xlsx'
Out<-'MADEP'

#data log
dataReceived[nrow(dataReceived)+1,]<-c(Out,TRUE,TRUE,NA)

#read field data 
  Field<-readField2015(In)
  dim(Field) #11 28

#read Lab data 
  Lab<-readLab2015(In)
  dim(Lab) #37 17

#fluorometerType changes (see above)
  Lab<-mutate(Lab,fluorometerType=replace(fluorometerType, analysisID%in%c(4:6), 'Other'))

#merge field and lab data
  assign(Out,full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate')))
  dim(get(Out))

#look for rows missing field or lab data 
    Chk<-filter(get(Out),is.na(orgID)|is.na(analysisID)) 
    if(nrow(Chk)==0){
      print("All records match")
        }else{
      select(Chk,waterbodyID, stationID,sampleID,sampleDate,analysisID)}#no lab data for Jamaica Pond and White Pond.  This is noted in the comments
```

### NHDES

* Monitoring data received from Sonya Carlson on 10/08/2015 (NH-for-EPA-2015-cyanoMonDataEntry.xlsx)
* Some additional monitoring data received from Sonya on 10/16/2015 (NH-for-EPA-2015-cyanoMonDataEntry-addenda.xlsx)
* A few more lines of monitoring data received from Sonya on 10/29/2015 (NH-for-EPA-2015-cyanoMonDataEntry-addenda2.xlsx)
* Secondary standards log received from Sonya on 11/13/2015 (NH-secondaryStandardsLog2015.xlsx)

**Message to Sonya sent 1/7/16 asking her to check the following Field and Lab corrections and to supply the missing field data**

* FieldDataEntry corrections
    * sampleID="BROOSSD02" changed to "BROOSD02" 
    * stationID="FORWCHEC02" changed to "FORWCHE02"
    * sampleID="FORWCHECO0201" changed to "FORWCHE0201"
    * sampleID="HALALTHWD01" changed to "HALALTHWB01"
    * stationID="SUNLNWL-GEN" changed to "SUNLNW-GEN"
    * waterbodyID="NHLAK801060402-05-01" changed to "NHLAK801060402-05-03"

* LabDataEntry corrections
  * sampleID for stationID="KEZSUTPR4" & sampleID="KEZSUTPIV401" changed to "KEZSUTPR401"
  * waterbodyID="NHLAK801060402-05-01" changed to "NHLAK801060402-05-03"

*missing field data
  * No field data for visit to waterbodyID="NHLAK700060401-02-02" sampleID="CRYEPA103" on 9/9/15
  * No field data for visit to waterbodyID="NHLAK700060401-02-02" sampleID="CRYEPA203" on 9/9/15
  * No field data for visit to waterbodyID="NHLAK700060401-02-02" sampleID="CRYGLMD03" on 9/9/15
  * No field data for visit to waterbodyID="NHLAK600020802-04-01" sampleID="OSSOSSD01" on 8/18/15
  * No field data for visit to waterbodyID="NHLAK700010804-02-01" sampleID="WEBEPA103" on 9/23/15
  * No field data for visit to waterbodyID="NHLAK700010804-02-01" sampleID="WEBEPA303" on 9/23/15

**Data steps**

* read and combine the 3 sets of monitoring data
* missing data and typos found (see above)



```{r NHDES, include=FALSE, echo=FALSE, cache=FALSE} 
#set data input and output names

Out<-'NHDES'

#data log
dataReceived[nrow(dataReceived)+1,]<-c(Out,TRUE,TRUE,"Missing Field and Lab Data")



#first set of data
In<-'NH-for-EPA-2015-cyanoMonDataEntry.xlsx'
#read field data 
  Field1<-readField2015(In)
  dim(Field1) #270 28

#read Lab data 
  Lab1<-readLab2015(In)
  dim(Lab1) #2115 17

#Second set of data
In<-'NH-for-EPA-2015-cyanoMonDataEntry-addenda.xlsx'
#read field data 
  Field2<-readField2015(In)
  dim(Field2) #1 28

#read Lab data 
  Lab2<-readLab2015(In)
  dim(Lab2) #9 17

#Third set of data
In<-'NH-for-EPA-2015-cyanoMonDataEntry-addenda2.xlsx'
#read field data 
  Field3<-readField2015(In)
  dim(Field3) #28 28

#read Lab data 
  Lab3<-readLab2015(In)
  dim(Lab3) #177 17

#combine field,field2 & field3
  Field<-rbind(Field1, Field2,Field3);nrow(Field) #299 (270+1+28)

#FieldDataEntry correction
  Field$sampleID[which(Field$sampleID=="BROOSSD02")]<-'BROOSD02' 
  Field$stationID[which(Field$stationID=="FORWCHEC02")]<-'FORWCHE02'
  Field$sampleID[which(Field$sampleID=="FORWCHECO0201")]<-'FORWCHE0201'
  Field$sampleID[which(Field$sampleID=="HALALTHWD01")]<-'HALALTHWB01'
  Field$stationID[which(Field$stationID=="SUNLNWL-GEN")]<-'SUNLNW-GEN'
  Field$waterbodyID[which(Field$waterbodyID=="NHLAK801060402-05-01")]<-'NHLAK801060402-05-03'

#combine Lab,Lab2 & Lab3
  Lab<-rbind(Lab1, Lab2,Lab3);nrow(Lab)#2301 (2115+177+9)

#LabDataEntry correction
  Lab$sampleID[which(Lab$stationID=="KEZSUTPR4" & Lab$sampleID=="KEZSUTPIV401")]<-'KEZSUTPR401'
  Lab$waterbodyID[which(Lab$waterbodyID=="NHLAK801060402-05-01")]<-'NHLAK801060402-05-03'

#merge field and lab data
  assign(Out,full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate')))
  dim(get(Out))

#look for rows missing field or lab data 
    Chk<-filter(get(Out),is.na(orgID)|is.na(analysisID)) 
    if(nrow(Chk)==0){
      print("All records match")
        }else{
      select(Chk,waterbodyID, stationID,sampleID,sampleDate,analysisID)}#6 rows with missing lab data and 82 rows with missing field data

#output file to xls to send back for revisions
    File<-paste('rawData/',Out,'dataToCheck.xls',sep='')
    write.xlsx(Chk, file=File,row.names=FALSE,showNA=FALSE)
#output flat file
    write.xlsx(NHDES, file='rawData/tempNHDES.xls',row.names=FALSE,showNA=FALSE)
```


### LIMTEX

* Monitoring data received from Nancy Leland on 11/15/2015 (Copy of cyanoMon2015DataEntry.xls)
* Secondary standards log received from Nancy Leland on 11/15/2015 (Copy of secondaryStandardsLog2015.xlsx)


**Data steps**

* read the data 
* sampleDepth entered as "3M"; changed to 3
* complete match between lab and field data


```{r LIMTEX, include=FALSE, echo=FALSE, cache=TRUE} 
#set data input and output names
In<-'Copy of cyanoMon2015DataEntry.xls'
Out<-'LIMTEX'

#data log
dataReceived[nrow(dataReceived)+1,]<-c(Out,TRUE,TRUE,NA)

#read field data 
  Field<-readField2015(In)
  dim(Field) #10 28

#update sampleDepth
Field$sampleDepthM<-3

#read Lab data 
  Lab<-readLab2015(In)
  dim(Lab) #20 17

#merge field and lab data
  assign(Out,full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate')))
  dim(get(Out))

#look for rows missing field or lab data 
    Chk<-filter(get(Out),is.na(orgID)|is.na(analysisID)) 
    if(nrow(Chk)==0){
      print("All records match")
        }else{
      select(Chk,waterbodyID, stationID,sampleID,sampleDate,analysisID)}#all good
```

### CTDEEP

* Monitoring data received from Tracy Lizotte on 12/10/2015 (cyanoMon2015DataEntry- CT DEEP.xls)
* A key to photos was received from Tracy Lizotte on 12/10/2015 (Cyanobacteria ID Legend 2015.xlsx)
    - NOTE: photos have not arrived yet but will
* Secondary standards log NOT received 
* Message sent to Tracy on 20160107 requesting she confirm the changes below
* fieldDataEntry corrections:
    * sampleID="18716-063015" changed to "18717-063015" for waterbodyID="18717
    * sampleID="18716-070715" changed to "18717-070715" for waterbodyID="18717
    * sampleID="18717-072115" changed to "18718-072115" for waterbodyID="18718
    * sampleID="18474-072115" changed to "18475-072115" for waterbodyID="18475
* labDataEntry Corrections:
    * sampleID=="18714-0817115" changed to "18714-081715"
    * sampleIDs: "18717-L-080315", "18717-M-080315", "18717-R-080315" changed to "18717-080315"
    * sampleDate: "2015-01-01" changed to "2015-08-11"
* 20160107: Tracy confirmed all changes-good to go.

**Data steps**

* read the data 
* waterbodyID and stationID are the same so I guess there is one station per lake
* some typos were corrected (see above)


```{r CTDEEP, include=FALSE, echo=FALSE, cache=TRUE} 
#set data input and output names
In<-'cyanoMon2015DataEntry- CT DEEP.xls'
Out<-'CTDEEP'

#data log
dataReceived[nrow(dataReceived)+1,]<-c(Out,TRUE,FALSE,"Some Missing Field and Lab Data")

#read field data 
  Field<-readField2015(In)
  dim(Field) #70 28

#Field data corrections
    Field$sampleID[which(Field$sampleID=="18716-063015" & Field$stationID=="18717")]<-"18717-063015"
    Field$sampleID[which(Field$sampleID=="18716-070715" & Field$stationID=="18717")]<-"18717-070715"
    Field$sampleID[which(Field$sampleID=="18717-072115" & Field$stationID=="18718")]<-"18718-072115"
    Field$sampleID[which(Field$sampleID=="18474-072115" & Field$stationID=="18475")]<-"18475-072115"

#read Lab data 
  Lab<-readLab2015(In)
  dim(Lab) #282 17

#Lab data corrections
    Lab$sampleID[which(Lab$sampleID=="18714-0817115")]<-"18714-081715"
    q<-which(Lab$stationID=="18717" & Lab$sampleDate=="2015-08-03") 
      Lab$commentAnalysis[q]<-paste("sampleID changed from",Lab$sampleID[q])
      Lab$sampleID[q]<-"18717-080315"
    Lab$sampleDate[which(Lab$sampleDate=="2015-01-01")]<-"2015-08-11"

#merge field and lab data
  assign(Out,full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate')))
  dim(get(Out))

#look for rows missing field or lab data 
    Chk<-filter(get(Out),is.na(orgID)|is.na(analysisID))
    ifelse(nrow(Chk)==0,"All records match",select(Chk,waterbodyID, stationID,sampleID,sampleDate,analysisID))

    if(nrow(Chk)==0){
      print("All records match")
        }else{
      select(Chk,waterbodyID, stationID,sampleID,sampleDate,analysisID)}#all good after corrections
```
  
### CFL

* Monitoring data received from Sabina Perkins on 12/29/2015 (CFL_cyanoMon2015DataEntry.xlsx)
* Secondary standards log not received 
* Sabina modified the Excel template so that all the data (Field and Lab) are in a single file.
* CFR used the older template with "parameter" & "valueUGL" instead of "ChlaUGL" & "PhycoUGL".
* For the phycocyanin readings there are a couple of samples that were read twice.  Change analysisRep for second reading to 'duplicate'.
* Lake "LBL" station "Off1", stationDescription shown as "Deep Site, NEAR Station 1" or "Deep Site, NEAR Station 2".  All changed to "Deep Site, NEAR Station 1"

**Data steps**

* read the data 
* make sure all fields are included
* check formatting
* add "ChlaUGL" & "PhycoUGL" fields


```{r CFL, include=FALSE, echo=FALSE, cache=FALSE} 
#set data input and output names
In<-'CFL_cyanoMon2015DataEntry.xlsx'
Out<-'CFL'

#data log
dataReceived[nrow(dataReceived)+1,]<-c(Out,TRUE,FALSE,NA)

#read the data
Data<-paste('rawData/',In,sep='') 
CFL<-read.xlsx(Data,sheetName='Field Data Entry',stringsAsFactors=FALSE)

#eliminate blank rows
CFL<-CFL[!is.na(CFL$waterbodyID),]

#check to make sure all variables are included
  #list of names for merged dataset
Names<-c('orgID','ContactName','Email','Phone','waterbodyID','waterbodyName','state','town','commentWB','stationID','stationDescription','stationType','longitudeSta','latitudeSta','locationSourceSta','commentStation','sampleID','sampleDate','sampleTime','fieldCrew','sampleMethod','sampleRep','sampleDepthM','waterTempC','photoSample','surfaceWaterCondition','weather','commentSample','analysisID','analysisDate','analystName','frozen','filtered','dilution','sampleTemperatureC','ChlaUGL','PhycoUGL','analysisRep','fluorometerType','photosAnalysis','commentAnalysis')

#check names
a<-names(CFL)%in%Names
  table(a)                      #FALSE==2 TRUE=#39
  names(CFL)[which(a==FALSE)]   #names(CFL) not in Names: "parameter" "valueUGL"

a<-Names%in%names(CFL)
  table(a)                      #FALSE==2 TRUE=#39
  Names[which(a==FALSE)]        #Names not it names(CFL): "ChlaUGL"  "PhycoUGL"

#add missing fields
Phyco<-filter(CFL,parameter=="Phycocyanin")
Phyco<-rename(Phyco,PhycoUGL=valueUGL)
nrow(Phyco) #168

Chla<-filter(CFL,parameter=="Chlorophyll")
Chla<-rename(Chla,ChlaUGL=valueUGL)
nrow(Chla) #165

#For the phycocyanin readings there are a couple of samples that were read twice.  Change second reading to duplicate.
Phyco$analysisRep[Phyco$waterbodyID=='LBL' & Phyco$stationID=='Off2' & Phyco$PhycoUGL==12.64 & Phyco$sampleDate=="2015-07-16"]<-'Duplicate'
Phyco$analysisRep[Phyco$waterbodyID=='WHP' & Phyco$stationID=='Off1' & Phyco$PhycoUGL==.1 & Phyco$sampleDate=="2015-07-23"]<-'Duplicate'
Chla$analysisRep<-as.character(Chla$analysisRep)

#update analysis comment so that it is shared by all samples
Phyco$commentAnalysis[Phyco$waterbodyID=='WHP' & Phyco$stationID=='Off1' & Phyco$sampleDate=="2015-07-23"]<-"Phycocyanin analysed twice because of an abnormally high value"

Chla$commentAnalysis[Chla$waterbodyID=='WHP' & Chla$stationID=='Off1' & Chla$sampleDate=="2015-07-23"]<-"Phycocyanin analysed twice because of an abnormally high value"

#update latitudeSta for one station
unique(Chla$latitudeSta[Chla$waterbodyID=='LBL' & Chla$stationID=='Off2'])  #all the same
unique(Phyco$latitudeSta[Phyco$waterbodyID=='LBL' & Phyco$stationID=='Off2']) #two values
Phyco$latitudeSta[Phyco$waterbodyID=='LBL' & Phyco$stationID=='Off2']<-unique(Chla$latitudeSta[Chla$waterbodyID=='LBL' & Chla$stationID=='Off2'])

#update stationDescription for LBL OFF2
distinct(select(filter(Phyco,waterbodyID=='LBL' & stationID=='Off1'),waterbodyID,stationID,stationDescription))
distinct(select(filter(Chla,waterbodyID=='LBL' & stationID=='Off1'),waterbodyID,stationID,stationDescription))
Chla$stationDescription[Chla$waterbodyID=='LBL' & Chla$stationID=='Off1']<-"Deep Site, NEAR Station 1"

#update "fieldCrew" for LBL OFF1 and Bant Off1
distinct(select(filter(Phyco,waterbodyID=='LBL' & stationID=='Off1'),waterbodyID,stationID,fieldCrew)) #one value
distinct(select(filter(Chla,waterbodyID=='LBL' & stationID=='Off1'),waterbodyID,stationID,fieldCrew))  #two values

distinct(select(filter(Phyco,waterbodyID=='Bant' & stationID=='Off1'),waterbodyID,stationID,fieldCrew)) #one value
distinct(select(filter(Chla,waterbodyID=='Bant' & stationID=='Off1'),waterbodyID,stationID,fieldCrew))  #one different value
  #change all to "Sabina Perkins, Elena Welch, Hillary Kenyon"
Chg<-"Sabina Perkins, Elena Welch, Hillary Kenyon"

Chla$fieldCrew[Chla$waterbodyID=='LBL' & Chla$stationID=='Off1']<-Chg
Chla$fieldCrew[Chla$waterbodyID=='Bant' & Chla$stationID=='Off1']<-Chg

Phyco$fieldCrew[Phyco$waterbodyID=='LBL' & Phyco$stationID=='Off1']<-Chg
Phyco$fieldCrew[Phyco$waterbodyID=='Bant' & Phyco$stationID=='Off1']<-Chg

#merge Phyco and Chla
CFL<-merge(Phyco,Chla,
           by=c('orgID','ContactName','Email','Phone','waterbodyID','waterbodyName','state','town','commentWB','stationID','stationDescription','stationType','longitudeSta','latitudeSta','locationSourceSta','commentStation','sampleID','sampleDate','sampleTime','fieldCrew','sampleMethod','sampleRep','sampleDepthM','waterTempC','photoSample','surfaceWaterCondition','weather','commentSample','analysisID','analysisDate','analystName','frozen','filtered','dilution','sampleTemperatureC','analysisRep','fluorometerType','photosAnalysis','commentAnalysis'),
           all=TRUE)


#reorder fields
  CFL<-CFL[,Names]

#reformat data
#character data
for(i in c(1:12,15:17,20:22,25:29,31:34,38:41)) CFL[,i]<-as.character(CFL[,i])
#numeric data
for(i in c(13,14,23,24,35:37)) CFL[,i]<-as.numeric(CFL[,i])
#date data
CFL$sampleDate<-as.Date(CFL$sampleDate)
CFL$analysisDate<-as.Date(CFL$analysisDate)
#time data (still needs work)
#CFL$sampleTime<-as.POSIXct(CFL$sampleTime)
```

### VTDEC

* Monitoring data received from Angela Shambaugh on 12/31/2015 (VTcyanoMon2015.xls)
* Secondary standards log not received but the data are included with the cyanoMon data.  Need to be extracted

# data steps
* read data
* add orgID='NHDES'
* update stationID, change lowercase "sta..." to uppercase "Sta...

# data corrections-these changes sent to Angela on 20160112 for confirmation-waiting reply
* field data changes
      * Field$stationID<-gsub('sta','Sta',x=Field$stationID)
      * Field$waterbodyID[Field$stationID=='Sta50']<-"VT05-01L01"

* lab data changes
      * Lab$stationID<-gsub('sta','Sta',x=Lab$stationID)
      * Lab$waterbodyID[Lab$stationID=="Sta04"]<-"VT04-02L01"
      * Lab$sampleID[Lab$sampleID==201542258]<-201542259
      * Lab$sampleDate[Lab$sampleID==201542103]<-"2015-05-06"
      * Lab$sampleID[Lab$sampleID==201542103]<-201542073
      * Lab$sampleID[Lab$sampleID==201542288]<-201542287

```{r VTDEC, include=FALSE, echo=FALSE, cache=FALSE} 
#set data input and output names
In<-'VTcyanoMon2015.xls'
Out<-'VTDEC'

#data log
dataReceived[nrow(dataReceived)+1,]<-c(Out,TRUE,FALSE,NA)

#read field data 
  Field<-readField2015(In)
  dim(Field) #65 28
  Field$orgID<-Out

#field data changes
  Field$stationID<-gsub('sta','Sta',x=Field$stationID)
  Field$waterbodyID[Field$stationID=='Sta50']<-"VT05-01L01"

#read Lab data 
  Lab<-readLab2015(In)
  dim(Lab) #152 17

#lab data changes
  Lab$stationID<-gsub('sta','Sta',x=Lab$stationID)
  Lab$waterbodyID[Lab$stationID=="Sta04"]<-"VT04-02L01"
  Lab$sampleID[Lab$sampleID==201542258]<-201542259
  Lab$sampleDate[Lab$sampleID==201542103]<-"2015-05-06"
  Lab$sampleID[Lab$sampleID==201542103]<-201542073
  Lab$sampleID[Lab$sampleID==201542288]<-201542287

#merge field and lab data
  assign(Out,full_join(Field, Lab, by=c('waterbodyID','stationID','sampleID','sampleDate')))
  dim(get(Out))#160 41

#look for rows missing field or lab data 
    Chk<-filter(get(Out),is.na(orgID)|is.na(analysisID)) 
    if(nrow(Chk)==0){
      print("All records match")
        }else{
      select(Chk,waterbodyID, stationID,sampleID,sampleDate,analysisID)}

#output file to xls to send back for revisions
    File<-paste('rawData/',Out,'dataToCheck.xls',sep='')
    #write.xlsx(Chk, file=File,row.names=FALSE,showNA=FALSE)
```

###Data merged saved as Data2015.rda and Data2015.xls 

```{r saveData, include=FALSE, echo=FALSE, cache=FALSE} 


Data2015<-rbind(CRWA,MADEP,NHDES,LIMTEX,CTDEEP,CFL,VTDEC)
write.xlsx(Data2015, file="Data2015.xls",row.names=FALSE,showNA=FALSE,sheetName="Data2015")
save(dataReceived,Data2015,file='Data2015.rda')
```


