---
title: "Locations2015"
author: "B"
date: "June 2, 2015"
output: html_document
---
<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  library(knitr)
    setwd("Data2015")
    knit('Locations2015.rmd')
  -->
```{r setup, include=FALSE, echo=FALSE, cache=FALSE} 
  #########function to install (if needed) and load R packages by list
libs<-c("rgdal","sp","knitr","maptools","rgeos","jsonlite","RODBC") #list of packages to load

installLoad<-function(pck){ #user defined function
    if(!pck%in%installed.packages()){install.packages(pck)}
    require(pck, character.only = TRUE)
  }
lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

To Do List
-------------------------
* Map locations and send to contacts for verifications

Introduction
-------------------------
This is the second year of the region 1 cyanobacteria monitoring program.  We are using the locations from last year as a starting point to populate the database with a few changes.

*  We received updated waterbodyID and stationdID information for NHDES from Sonya Carlson on 5/21/15. The name of the file is: "2015-Probable-NH-Waterbody-IDs-StationIDs.xlsx"
*  The prefixes of the following waterbodyIDs were changed as follows
    - MEDEP:  prefix "MIDAS_" was replaced with "MEDEP_"
    - UNH: prefix "UNH_CFB_WBID_" was replaced with "UNH_"
    - VTDEC: prefix "VTDEC_WBID_" was replaced with "VTDEC_"
    

Data Steps
-------------------------

* Download the data 
* For data definitions see: https://github.com/willbmisled/Reg1Cyano/blob/master/Access/cyanoMon2015Documentation.md
* write data to xlsx file

```{r Data, include=FALSE, echo=FALSE, cache=FALSE} 
#get locations from the cyanoMon2015.mdb   
  con <- odbcConnectAccess("C:/Users/WMilstea/SharePoint/Warren, Phil/NLA/cyanoMon2015.mdb")
    Loc <- sqlQuery(con, "
SELECT tblWaterbody.waterbodyID, tblWaterbody.waterbodyName, tblStation.stationID, tblStation.stationDescription, tblStation.stationType, tblWaterbody.state, tblWaterbody.town, tblStation.longitudeSta, tblStation.latitudeSta, tblStation.locationSourceSta, tblStation.commentStation
FROM tblWaterbody LEFT JOIN tblStation ON tblWaterbody.waterbodyID = tblStation.waterbodyID
ORDER BY tblWaterbody.waterbodyID, tblStation.stationID;
",stringsAsFactors=FALSE)
close(con)
str(Loc)
head(Loc)

#write xls
  xlsFile <- odbcConnectExcel("Locations2015.xls", readOnly = FALSE)
  sqlSave(xlsFile,Loc, append=FALSE,rownames=FALSE)
  odbcCloseAll()
```

* Remove missing values

* Convert locations to Spatial Points Dataframe

```{r spdf, include=FALSE, echo=FALSE, cache=FALSE} 
  #remove missing sites with missing location data
    a<-Loc[!is.na(Loc$longitudeSta),]

#create spatial points dataframe
  WGS84<-CRS("+proj=longlat +datum=WGS84")  #ESRI GCS_WGS_1984 
  Loc1<-SpatialPointsDataFrame(coordinates(a[,c("longitudeSta","latitudeSta")]), a, proj4string=WGS84)
plot(Loc1)
```

* create a combined waterbodyID stationID name for KML index
* create a description from waterbodyName, stationDescription, and commentStation
* create KML and shapefile of spatial data

```{r KML, include=FALSE, echo=FALSE, cache=FALSE}
  kmlID<-paste(a$waterbodyID,"_",a$stationID,sep='')
  kmlDesc<-paste("Name = ",a$waterbodyName,"; Location = ",a$stationDescription, "; Comment = ",a$commentStation,sep='')
#create KML file of location          
  kmlPoints(Loc1, 
            kmlfile='Locations2015.kml', 
            name=kmlID, 
            description=kmlDesc,
            icon="http://maps.google.com/mapfiles/kml/paddle/ltblu-stars.png",
            kmlname="Locations2015",
            kmldescription="2015 Region 1 CyanoMonitoring Locations")
#open file in google earth
  #
shell.exec('Locations2015.kml')


#save shapefile
writeOGR(Loc1,getwd(),'Locations2015', driver="ESRI Shapefile",overwrite_layer=TRUE)
  
```

* create json files of waterbodyIDs and stationIDs for the phone app
* the structure of waterBodyIDList.json will be:
    - {“waterBodyID":[“wbid1”,"wbid2”,”wbid3”,”wbid4”,”wbid5”]}
* the structure of stationIDList.json will be:
    - {“stations":{“wbid1":[“sta1”,"sta2”,"sta3"]
    - ,”wbid2":[“sta1”,"sta2”,"sta3”]
    - ,”wbid3":[“sta1”,"sta2”,"sta3”]
    - ,”wbid4":[“sta1”,"sta2”,"sta3”]
    - ,”wbid5":[“sta1”,"sta2”,"sta3”]
    - }}


```{r json, include=FALSE, echo=FALSE, cache=FALSE}
#waterbodyIDs
  waterbodyIDList<-toJSON(list(waterbodyID=unique(Loc$waterbodyID)))

#stationIDs
  #get waterbodyIDs
    w<-unique(Loc$waterbodyID)
  #create lists by waterbodyID
    #start list
      a<-list(stations=list(NA))
    #loop through waterbodyIDs to create big list
      #for(i in 1:length(w)){
      for(i in 1:length(w)) a$stations[i]<-list(Loc$stationID[Loc$waterbodyID==w[i]])
    #rename lists
       names(a$stations)<-w        
  #create JSON
    stationIDList<-toJSON(a)

#output JSON files
  write(stationIDList, "stationIDList.json")
  write(waterbodyIDList, "waterbodyIDList.json")

####NOTE: both JSON files validated here:  http://jsonlint.com/
```

