---
title: "readJSON"
author: "Bryan Milstead"
date: "Wednesday, November 05, 2014"
output: html_document
---
<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  library(knitr)
      setwd('Data2015/json')
  knit('readJSON.rmd')
-->

To Do List
-------------------------
* C


Introduction
-------------------------

This is a

```{r setup, include=FALSE, echo=FALSE, cache=FALSE} 
  #change the working directory
    tryCatch(setwd('Data2015/json'),error = function(e) NA)
  #load libraries
    library(rjson)
```


```{r readJSON, include=FALSE, echo=FALSE, cache=FALSE} 
options(stringsAsFactors=FALSE)
#Switch to directory where the JSON files are located
  list.dirs()
  Dir<-"NancyLeland/20150806"
  setwd(Dir)

#get a list of the files and sort by type: crew; waterbody; station; sample; analysis;
  b<-list.files(pattern='.json');b
  crew<-b[grep('CREW.json',b)]
  waterbody<-b[grep('WATERBODY',b)]
  station<-b[grep('STATION',b)]
  sample<-b[grep('SAMPLE',b)]
  analysis<-b[grep('ANALYSIS',b)]

#format org data
  org<-as.data.frame(fromJSON(file=crew)[c(4,1,2,3)])
  names(org)<-c('orgID','contact','email','phone')

#format waterbody data
  wb<-as.data.frame(fromJSON(file=waterbody))
  rename(wb,commentWB=commentWaterbody)#rename to match database
  wb$orgID<-org$orgID #add the orgID NOTE: we are assuming that only one orgID is contributing to this directory

#format station data 
  sta<-c()
  if(length(station)>0) {
    for(i in (1:length(station))){
      b<-cbind(as.data.frame(fromJSON(file=station[i])))
      sta<-rbind(sta,b)
    }
    #add the waterbodyID
      sta$stationID<-gsub("\\#","_",sta$stationID) #eliminate the special character "x" from the analysis ID
      sta$waterbodyID<-t(as.data.frame(sapply(sta$stationID,strsplit,"_")))[,1]
}                    
                    
                  
 

#format sample data 
sa<-c()
if(length(sample)>0) {
  for(i in (1:length(sample))){
    b<-cbind(as.data.frame(fromJSON(file=sample[i])))
    sa<-rbind(sa,b)
  }
  #add the stationID
      sa$sampleID<-gsub("\\#","_",sa$sampleID) #eliminate the special character "x" from the analysis ID
      q<-t(as.data.frame(sapply(sa$sampleID,strsplit,"_"))) #parse the analysis ID
      sa$stationID<-paste(q[,1],'_',q[,2],sep='')
      rename(sa,sampleDate=date)#rename to match database
}  

#format analysis data 
  an<-c()
  if(length(analysis)>0) {
    for(i in (1:length(analysis))){
      b<-cbind(as.data.frame(fromJSON(file=analysis[i])))
      an<-rbind(an,b)
    }
  #parse waterbdodyID, stationID and sampleID
    an$analysisID<-gsub("\\#","_",an$analysisID) #eliminate the special character "x" from the analysis ID
    q<-as.data.frame(t(as.data.frame(sapply(an$analysisID,strsplit,"_")))) #parse the analysis ID
    an$stationID<-paste(q[,1],'_',q[,2],sep='')
    an$sampleID<-paste(an$stationID,'_',q[,3],sep='')
  #add sampleDate
    q<-merge(an,sa,by='sampleID',all.x=TRUE)
    }



merge(st,sa,by='stationID')





```  
    			

                      
                      
                    
  
  




grep('ANALYSIS',b)

i<-1;fromJSON(file=b[i])

i<-2

a<-fromJSON(file=b[i])
Names<-names(a)




a[names(a)[1]]







b<-c("Lake#Champlain#0601#649_WATERBODY.json",
     "Lake#Champlain#0601#649#ST1_STATION.json",
     "Lake#Champlain#0601#649#ST1#SAMP1#ANL1_ANALYSIS.json",
     "Lake#Champlain#0601#649#ST1#SAMP1_SAMPLE.json",       
     "Lake#Champlain#0601#649#ST1#SAMP3_SAMPLE.json")   

for(i in 1:4) str(fromJSON(file=b[i]))

i=1;str(fromJSON(file=b[i]));fromJSON(file=b[i])

str(fromJSON(file=b["3814_NAME_submission.json"]))


str(fromJSON(file="#ST1_STATION.json"))
 

```

