---
title: "Region1CyanoLocations2014"
author: "Bryan Milstead"
date: "January 14, 2015"
output: html_document
---
<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  library(knitr)
      a<-getwd()
        if(substr(a,nchar(a)-8,nchar(a))=='Reg1Cyano')  {setwd('./Data2014/')
          } else {if(substr(a,nchar(a)-7,nchar(a))!='Reg1Cyano') print('WARNING: Wrong Working Directory')}
  knit('Locations2014.rmd')
  -->
```{r setup, include=FALSE, echo=FALSE, cache=FALSE} 
  #########function to install (if needed) and load R packages by list
libs<-c("rgdal","sp","knitr","maptools","rgeos") #list of packages to load

installLoad<-function(pck){ #user defined function
    if(!pck%in%installed.packages()){install.packages(pck)}
    require(pck, character.only = TRUE)
  }
lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

To Do List
-------------------------
* missing location data for ME
* check locations manually for distance >0
* Map locations and send to contacts for verifications

Introduction
-------------------------
During the summer of 2014 the New England states initiated a monitoring program for cyanobacteria in lakes.  Participants included state and tribal governments, local watershed associations, EPA Region 1, and the EPA Atlantic Ecology Division. 

The data have been standardized and are available here in .csv format (hint: right click on the link and choose “Save As” to save the file to your computer, otherwise, it will open in your browser): https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Data2014.csv?raw=true

The details of the data processing steps can be found here: https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Data2014.md

This document is available here: https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Locations2014.md

A KML file for view in Google Earth is available here (hint: right click on the link and choose “Save As” to save the file to your computer, otherwise, it will open in your browser): https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Locations.kml?raw=true

An R datafile with a the Locations in SpatialPointsDataframe format is available here (hint: right click on the link and choose “Save As” to save the file to your computer, otherwise, it will open in your browser): https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Locations.rda?raw=true

Data Definitions SpatialPointsDataFrame "Locations" in "Locations.rda" `r nrow(Locations)` observations
-------------------------
Field  | Units | Description
------------- | ------------- | -------------
**Longitude:**|(Decimal Degrees)|Longitude of Location
**Latitude:**|(Decimal Degrees)|Latitude of Location
**Location:**|(Integer)|Unique identifier for location
**WBID:**|(Integer)|Unique identifier for closest lake to location
**Distance:**|(Meters)|Distance from Location to nearest lake

Data Sources
-------------------------
The field data, including location information, have been standardized and are available here in .csv format (hint: right click on the link and choose “Save As” to save the file to your computer, otherwise, it will open in your browser): https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Data2014.csv?raw=true

The details of the data processing steps and the data definitions can be found here: https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Data2014.md

This document is available here: https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Locations2014.md

Data Steps
-------------------------

* Download the data `r Data2014<-read.csv("Data2014.csv")`
* Check which data (excluding blanks and standards) are missing locations
* Created NHDES_MissingLocation.csv & UNH_MissingLocation.csv

```{r missing, include=FALSE, echo=FALSE, cache=FALSE} 
#select samples only
  unique(Data2014$SampleMethod)
  a<-Data2014[Data2014$SampleMethod=='Grab'|Data2014$SampleMethod=='VanDorn'|Data2014$SampleMethod=='Integrated',]
#which locations are missing
  table(is.na(a$Latitude)) #0 missing
```

* Keep only the unique long lat data 
* Add a unique identifier "Location" 
* Create a SpatialPointsDataFrame

```{r spdf, include=FALSE, echo=FALSE, cache=FALSE} 
 #Download the data
Loc<-unique(na.exclude(Data2014[,c("Longitude","Latitude")])) #Keep only the unique long lat data
Loc$Location<-c(1:nrow(Loc)) #unique identifier

#create spatial points dataframe
  WGS84<-CRS("+proj=longlat +datum=WGS84")  #ESRI GCS_WGS_1984 
  Loc<-SpatialPointsDataFrame(coordinates(Loc[,-3]), Loc, proj4string=WGS84)
plot(Loc)

#transform to Albers
  #ESRI USA Contiguous Albers Equal Area Conic (used by MRB1 WBIDLakes as AlbersX and AlbersY)
  AlbersContiguous<-CRS('+proj=aea +x_0=0 +y_0=0 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +units=m +datum=NAD83')
    LocA<-spTransform(Loc,AlbersContiguous) #reproject
```

* Get MRB1_WBIDLakes from the WaterbodyDatabase.mdb
* Reproject to WGS84
* Fix holes (just in case)
```{r getLakes, include=FALSE, echo=FALSE, cache=TRUE} 
library(rgdal)
library(maptools)
  MDB<-"c:/bryan/EPA/Data/WaterbodyDatabase/WaterbodyDatabase.mdb" #data source
  features <- ogrListLayers(dsn=MDB) #MRB1_WBIDLakes
  Lakes<-readOGR(dsn=MDB,'MRB1_WBIDLakes')
  Lakes<-spTransform(Lakes,WGS84) #reproject
  #Associate lake holes (islands) with the correct polygon.  This is an important step.
    slot(Lakes, "polygons") <- lapply(slot(Lakes, "polygons"), checkPolygonsHoles)

#transform to Albers
     LakesA<-spTransform(Lakes,AlbersContiguous) #reproject
```

* Find the closest lake to each point
* First buffer the points by 1000 meters
* Use Over to select lakes within the buffer
* Use gDistance to find closest lake to each point
* If no lakes are found iteratively, increase the buffer distance up to 50km.

```{r gDist, include=FALSE, echo=FALSE, cache=FALSE} 
#buffer points and select lakes within the buffer.
  #dataframe to hold results
    WBID<-data.frame(WBID=rep(NA,nrow(LocA)),Distance=rep(NA,nrow(LocA)))  
  #loop through the point locations
    for(i in 1:nrow(LocA)) {  
      #select Point
        pt<-LocA[i,]
      #While Loop to select Lakes  
        BufLakes<-data.frame() #reset BufLakes
        D<-1000 #reset buffer distance
        #start Loop
         while(nrow(BufLakes)==0) {
          if (D>50001) break #stop if D greater than 50km
          #Buffer Point
            Buf<-gBuffer(pt,width=D)
          #increase buffer width for next run
            D<-D+10000 
          #Select Lakes within Buffer-returns empty data.frame if no lakes are selected
            BufLakes<-tryCatch(LakesA[unlist(over(Buf, LakesA, returnList = TRUE)),],error = function(e) data.frame())    
          }      
      # find closest BufLake to point with gDistance
        if(nrow(BufLakes)>0){
            a<-vector() #empty work vector
          for(j in 1:nrow(BufLakes)) {  #loop through the BufLakes
            a<- append(a, gDistance(LocA[i,],BufLakes[j,])) #vector with distance from pt to each lake
            Dist<-min(a) #distance to closest lake
            WB<-BufLakes@data$WB_ID[which(a==Dist)] #WBID for closest lake
            b<-data.frame(WBID=WB,Distance=Dist) #output
          }
        }else{
            b<-data.frame(WBID=NA,Distance=NA) #output = NA
        }
        WBID[i,]<-b #add output to results df
    } 

  #add to Loc and LocA
    Loc@data$WBID<-WBID$WBID
    Loc@data$Distance<-WBID$Distance
    LocA@data$WBID<-WBID$WBID
    LocA@data$Distance<-WBID$Distance

```

* Create KML file of locations

```{r KML, include=FALSE, echo=FALSE, cache=FALSE} 
#create KML file of location          
  kmlPoints(Loc, 
            kmlfile='Locations.kml', 
            name=paste("Location",Loc@data$Location), 
            description=paste("WBID",Loc@data$WBID),
            icon="http://google.com/mapfiles/kml/paddle/wht-diamond.png",
            kmlname="Locations",
            kmldescription="Region 1 CyanoMonitoring Locations")
#open file in google earth
  #shell.exec('Locations.kml')
```

* rename SpatialPointsDataFrame Loc to Locations `r Locations<-Loc'
* save Locations as SpatialPointsDataFrame and SpatialPolygonsDataFrame "Lakes" 

```{r Save, include=FALSE, echo=FALSE, cache=FALSE} 
  save(Locations,Lakes,file='Locations.rda')
```





