---
title: "buildAccessDB"
author: "Bryan"
date: "April 09, 2015"
output: html_document
---

<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  library(knitr)
    setwd("Access")
    knit('buildAccessDB.rmd')
  -->
  
```{r setup, include=FALSE, echo=FALSE, cache=FALSE} 
  #########function to install (if needed) and load R packages by list
libs<-c("knitr","chron") #list of packages to load

installLoad<-function(pck){ #user defined function
    if(!pck%in%installed.packages()){install.packages(pck)}
    require(pck, character.only = TRUE)
  }
lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

To Do List
-------------------------

* Develop a method to assign unique identifiers that works with the phone app
* Build forms
* Add 2014 data
* Develop and test phone app
* reassign values for stationType (tblStation) to "nearShore" and "offShore"

Background
-------------------------

* EPA region 1 is coordinating a Cyanobacteria Monitoring Progam for the six New England States (CT, MA, ME, NH, RI, & VT)
* Data collection initiated during the summer of 2014
* 2014 data have been collated and standardized: https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Data2014.md
* 2014 locations have been verified: https://github.com/willbmisled/Reg1Cyano/blob/master/Data2014/Locations2014.md
* A MSaccess database has been developed: https://github.com/willbmisled/Reg1Cyano/blob/master/Access/cyanoMonDocumentation.md
* OBJECTIVE: add the 2014 locations and data to cyanoMon.mdb

Datasteps
-------------------------

* Get data
* Keep samples only (eliminate blanks and standards)
* Merge tablular (df "Data") and location (df "LocID2LocIDNEW") data for Lakes
* Check that all data have locations-yes
* Rename fields to match cyanoMon.mdb (Old:New)
    - WaterbodyName:waterbodyName
    - State:state
    - Lon_DataID:longitudeData
    - Lat_DataID:latitudeData
    - Lon_LocIDNew:longitudeSta
    - Lat_LocIDNew:latitudeSta
    - GPSType:locationSourceSta
    - SampleLocation:stationType

```{r Data, include=FALSE, echo=FALSE, cache=FALSE} 
#reset working directory
  setwd("~/PortableApps/R/scripts/Reg1Cyano/Access")
  #setwd("C:/Bryan/PortableApps/R/scripts/Reg1Cyano/Access")
#Get the data 
  Data2014<-read.csv("../Data2014/Data2014.csv")
#Get the locations
 load("../Data2014/Locations.rda")
#Get the NH waterbodyNames
  NH<-read.csv("../Data2014/NHDES_waterbodyNames.csv") #get names NHDES
#select samples only
  unique(Data2014$SampleMethod)
  Data<-Data2014[Data2014$SampleMethod=='Grab'|Data2014$SampleMethod=='VanDorn'|Data2014$SampleMethod=='Integrated',]
  nrow(Data) #8499

#merge tabular data with locations
      Data<-merge(Data,LocID2LocIDNew,by='LocID')
        nrow(Data) #8499

#make sure all data have locations
      table(is.na(Data$Lon_LocIDNew)) #0 missing  N=8499
      table(is.na(Data$Lon_WBID)) #0 missing  N=8499

#rename fields
  names(Data)[which(names(Data)=="WaterbodyName")]<-"waterbodyName"
  names(Data)[which(names(Data)=="State")]<-"state"
  names(Data)[which(names(Data)=="Lon_WBID")]<-"longitudeWB"
  names(Data)[which(names(Data)=="Lat_WBID")]<-"latitudeWB"
  names(Data)[which(names(Data)=="Lon_LocIDNew")]<-"longitudeSta"
  names(Data)[which(names(Data)=="Lat_LocIDNew")]<-"latitudeSta"
  names(Data)[which(names(Data)=="GPSType")]<-"locationSourceSta"
  names(Data)[which(names(Data)=="SampleLocation")]<-"stationType"

  
```

**build tblWaterbody**

* Assign a unique waterbodyID to each lake-this will be the primary Key for tblWaterbody. Varies by Organization:
    - CRWA, CTDEEP, RIWW,VTDEC: Organization + SiteID
    - MEDEP: "MIDAS" + SiteID
    - NHDES: Organization + WaterbodyName
    - UNH_CFB & VTDEC: "WBID" + WBID
* Assign a otherWaterbodyID to each lake. Varies by Organization: 
    - CRWA, CTDEEP, RIWW,VTDEC: SiteID
    - VTDEC, UNH_CFB: NA #no ID given
    - NHDES: Organization + WaterbodyName #waterbody names are numeric IDs
* Make sure all waterbodyNames are consistently spelled
    - MIDAS_5408  Webber==Weber pond
    - MIDAS_5172  Unity==Unity Pond
* add missing lake names
    - VTDEC: lake champlain
    - NHDES: names manually extracted from "NHDES_Locations.csv" saved as "NHDES_waterbodyNames.csv" then added to DF "WB"
* add town==NA as a placeholder
* add field "locationSourceWB"
    - No WBID=='Google Earth
    - With WBID =='MRB1_WBIDLakes'
* extract commentWB and commentSta for siteComment
* Get unique values for c("waterbodyID","waterbodyName","state","town","WBID","otherWaterbodyID","LongitudeWB","LatitudeWB","locationSourceWB","commentWB")
* check to make sure the waterbodyID is unique and consistent

```{r tblWaterbody, include=FALSE, echo=FALSE, cache=FALSE}   
#assign otherWaterbodyID
  Data$otherWaterbodyID<-as.character(Data$SiteID)
    x<-Data$Organization=='UNH_CFB'|Data$Organization=='VTDEC'
      Data$otherWaterbodyID[x]<-NA
    x<-Data$Organization=='NHDES'
      Data$otherWaterbodyID[x]<-paste('NHDES','_',Data$waterbodyName[x],sep='')
  unique(Data$otherWaterbodyID)

#assign waterbodyID
  Data$waterbodyID<-paste(Data$Organization,'_',Data$SiteID,sep='')
    x<-Data$Organization=='MEDEP'
      Data$waterbodyID[x]<-paste('MIDAS','_',Data$SiteID[x],sep='')
    x<-Data$Organization=='UNH_CFB'|Data$Organization=='VTDEC'
      Data$waterbodyID[x]<-paste(Data$Organization[x],'_','WBID','_',Data$WBID[x],sep='')
    x<-Data$Organization=='NHDES'
      Data$waterbodyID[x]<-paste('NHDES','_',Data$waterbodyName[x],sep='')
        #remove spaces from one ID
          x<-grep('MA70-02',Data$waterbodyID)
            Data$waterbodyID[x]<-"CRWA_MA70-02"
            Data$otherWaterbodyID[1]<-"MA70-02"
  unique(Data$waterbodyID)

#waterbodyName consistency check  
  a<-unique(Data[,c('waterbodyID','waterbodyName')])
    x<-as.data.frame(table(a$waterbodyID))
      b<-x[x$Freq>1,]
        a[a$waterbodyID%in%b$Var1,]
    #waterbodyID with multiple lines due to spelling differences
        Data$waterbodyName[which(Data$waterbodyName=='Unity')]<-'Unity Pond'
        Data$waterbodyName[which(Data$waterbodyName=='Webber')]<-'Webber pond'

#missing or incorrect waterbodyName
  # add new levels to factor (Data$waterbodyName)
     levels(Data$waterbodyName)<-c(levels(Data$waterbodyName),'Lake Champlain',levels(NH$waterbodyName))
    #VTDEC
      Data$waterbodyName[grep("VTDEC",Data$waterbodyID)]<-"Lake Champlain"
    #NHDES
      Data<-merge(Data,NH,by='waterbodyID',all.x=TRUE)
        names(Data)[which(names(Data)=="waterbodyName.x")]<-"waterbodyName"
          x<-grep("NHDES",Data$waterbodyID)
            Data$waterbodyName[x]<-Data$waterbodyName.y[x]
#add town==NA
  Data$town<-NA

#add locationSourceWB
  Data$locationSourceWB<-ifelse(is.na(Data$WBID),'GoogleEarth','MRB1_WBIDLakes')

# extract commentWB and commentSta for siteComment
  # check comments for waterbody and station
    unique(Data$siteComment)
    #commentWB
      Data$commentWB<-NA
        x<-grep('WBID',Data$siteComment) #only a few wb comments
          Data$commentWB[x]<-as.character(Data$siteComment[x])
    #commentSta
      Data$commentSta<-Data$siteComment

#get unique values for tblWaterbody
    tblWaterbody<-unique(Data[,c("waterbodyID","waterbodyName","state","town","WBID","otherWaterbodyID","longitudeWB","latitudeWB","locationSourceWB","commentWB")]);nrow(tblWaterbody)#106
    head(tblWaterbody)

#waterbodyID consistency check  
  #repeated waterbodyIDs
    a<-as.data.frame(table(tblWaterbody$waterbodyID))
      b<-a[a$Freq>1,]
        tblWaterbody[tblWaterbody$waterbodyID%in%b$Var1,]
    #waterbodyID with multiple lines due to spelling differences
        tblWaterbody$waterbodyName[which(tblWaterbody$waterbodyName=='Unity')]<-'Unity Pond'
        tblWaterbody$waterbodyName[which(tblWaterbody$waterbodyName=='Webber')]<-'Webber pond'
```

**build tblStation**

* stationID: unique identifier for tblStation-combination of LocIDNew and stationType 
* Assign otherStationID to each station within each lake. Varies by Organization:
    - CRWA,MEDEP:  SiteLocation
    - CTDEEP,RIWW,UNH_CFB,: NA
    - NHDES,VTDEC: SiteID
* stationLocation-varies by organization
    - CRWA,CTDEEP,RIWW,UNH_CFB,NHDES:  siteLocation
    - MEDEP,VTDEC: NA
* stationType: field "sampleLocation" renamed "stationType" above Okay for now but will need to reassign values to "nearShore" and "offShore"
* Get unique values for c("stationID","waterbodyID","otherStationID","stationLocation","stationType","LongitudeSta","LatitudeSta","locationSourceSta","commentSta")
* check to make sure the waterbodyID is unique and consistent


```{r tblStation, include=FALSE, echo=FALSE, cache=FALSE}
#Assign otherStationID
  Data$otherStationID<-NA
    x<-Data$Organization=='CRWA'|Data$Organization=='MEDEP'
      Data$otherStationID[x]<-as.character(Data$SiteLocation[x])
    x<-Data$Organization=='NHDES'|Data$Organization=='VTDEC'
      Data$otherStationID[x]<-as.character(Data$SiteID[x])
  table(Data$Organization,is.na(Data$otherStationID))
  table(Data$Organization,Data$otherStationID)

#stationLocation
  Data$stationLocation<-Data$SiteLocation
    x<-Data$Organization=='VTDEC'|Data$Organization=='MEDEP'
      Data$stationLocation[x]<-NA
    table(Data$Organization,is.na(Data$stationLocation))

#Assign stationID
  Data$stationID<-paste(Data$LocIDNew,"_",Data$stationType,sep='')
  #modify stationID for LocIDNew==158
    x<-Data$LocIDNew==158
      Data$stationID[x]<-paste(Data$LocIDNew[x],"_",Data$stationType[x],"_",Data$stationLocation[x],sep='')
        Data$stationID[x]<-gsub(" ","",Data$stationID[x])
  #update locationSorceSta for LocIDNew==171 or LocIDNew==206
    x<-Data$LocIDNew==171|Data$LocIDNew==206
      Data$locationSourceSta[x]<-NA

#get unique values for tblStation
    tblStation<-unique(Data[,c("stationID","waterbodyID","otherStationID","stationLocation","stationType","longitudeSta","latitudeSta","locationSourceSta","commentSta")]);nrow(tblStation)#468

#stationID consistency check  
  #repeated stationIDs
    a<-as.data.frame(table(tblStation$stationID))
      b<-a[a$Freq>1,]
        x<-tblStation[tblStation$stationID%in%b$Var1,]
        x[order(x$stationID),]
```

**build tblSample**

* add formatted sample and analysis dates and times.
* Rename fields to match cyanoMon.mdb (Old:New)
    - Organization:organization
    - NameOfSamplers:fieldCrew
    - SampleMethod:sampleMethod
    - Depth:sampleDepthM
    - WeatherConditions:weather
    - Photos:photoSample
* Add placeholder fields
    - waterTempC
    - surfaceWaterCondition
    - commentSample
* Get unique values for c("stationID","sampleDate","sampleTime","organization","fieldCrew","sampleMethod","sampleDepthM","waterTempC","weather","surfaceWaterCondition","photoSample","commentSample")
* sampleID: create sampleID as 1:nrow(tblSample)
* check to make sure the sampleID is unique and consistent
* add sampleID to df Data so it can used in the next table
 


```{r tblSample, include=FALSE, echo=FALSE, cache=FALSE}
#add formatted date
  Data$sampleDate<-dates(paste(Data$SampleMonth,'/',Data$SampleDay,'/',Data$SampleYear,sep=''))
    unique(Data[,c('SampleMonth','SampleDay','SampleYear','sampleDate')])
  Data$analysisDate<-dates(paste(Data$AnalysisMonth,'/',Data$AnalysisDay,'/',Data$AnalysisYear,sep=''))
    unique(Data[,c('AnalysisMonth','AnalysisDay','AnalysisYear','analysisDate')])

#add formatted time
  Data$sampleTime<-times(paste(Data$SampleHour,':',Data$SampleMinutes,":00",sep=''))
    unique(Data[,c('SampleHour','SampleMinutes','sampleTime')])
  Data$analysisTime<-times(paste(Data$AnalysisHour,':',Data$AnalysisMinutes,":00",sep=''))
    unique(Data[,c('AnalysisHour','AnalysisMinutes','analysisTime')])

#rename fields
  names(Data)[which(names(Data)=="Organization")]<-"organization"
  names(Data)[which(names(Data)=="NameOfSamplers")]<-"fieldCrew"
  names(Data)[which(names(Data)=="SampleMethod")]<-"sampleMethod"
  names(Data)[which(names(Data)=="Depth")]<-"sampleDepthM"
  names(Data)[which(names(Data)=="WeatherConditions")]<-"weather"
  names(Data)[which(names(Data)=="Photos")]<-"photoSample"

#add placeholder fields
  Data$waterTempC<-NA
  Data$surfaceWaterCondition<-NA
  Data$commentSample<-NA

#create tblSample
  tblSample<-unique(Data[,c("stationID","sampleDate","sampleTime","organization","fieldCrew","sampleMethod","sampleDepthM","waterTempC","weather","surfaceWaterCondition","photoSample","commentSample")]);nrow(tblSample) #1102

#add sampleID
  sampleID<-1:nrow(tblSample)
  tblSample<-data.frame(sampleID,tblSample)

#check to make sure the sampleID is unique and consistent
  #repeated sampleIDs
      a<-as.data.frame(table(tblSample$sampleID))
        b<-a[a$Freq>1,]
          x<-tblSample[tblSample$sampleID%in%b$Var1,]
          x[order(x$sampleID),]

#add sampleID to df Data
  #select "by" fields
    #all
      nrow(unique(tblSample[,c("stationID","sampleDate","sampleTime","organization","fieldCrew","sampleMethod","sampleDepthM","weather","photoSample")])) #1102
    #subset
      nrow(unique(tblSample[,c("stationID","sampleDate","sampleTime","sampleMethod","sampleDepthM","weather")])) #1102
  #merge to df Data
    a<-merge(Data,tblSample,by=c("stationID","sampleDate","sampleTime","sampleMethod","sampleDepthM","weather"))
    Data<-merge(Data,a[,c('ID','sampleID')],by='ID')
```

**build tblAnalysis**

* Rename fields to match cyanoMon.mdb (Old:New)
    - Frozen:frozen
    - Filtered:filtered
    
* Add placeholder fields
    - analystName
    - dilution
    - sampleTemperatureC
    - photosAnalysis
    - commentAnalysis
* Get unique values for c('sampleID','analysisDate','analysisTime','analystName','frozen','filtered','dilution','sampleTemperatureC','photosAnalysis','commentAnalysis')

```{r tblAnalysis, include=FALSE, echo=FALSE, cache=FALSE}
#rename fields
  names(Data)[which(names(Data)=="Frozen")]<-"frozen"
  names(Data)[which(names(Data)=="Filtered")]<-"filtered"
  
#add placeholder fields
  Data$analystName<-NA
  Data$dilution<-NA
  Data$sampleTemperatureC<-NA
  Data$photosAnalysis<-NA
  Data$commentAnalysis<-NA

#create tblAnalysis
  tblAnalysis<-unique(Data[,c('sampleID','analysisDate','analysisTime','analystName','frozen','filtered','dilution','sampleTemperatureC','photosAnalysis','commentAnalysis')]);nrow(tblAnalysis) #2458

#add analysisID
  analysisID<-1:nrow(tblAnalysis)
  tblAnalysis<-data.frame(analysisID,tblAnalysis)

#check to make sure the analysisID is unique and consistent
  #repeated analysisIDs
      a<-as.data.frame(table(tblAnalysis$analysisID))
        b<-a[a$Freq>1,]
          x<-tblAnalysis[tblAnalysis$analysisID%in%b$Var1,]
          x[order(x$analysisID),]

#add analysisID to df Data
  #select "by" fields
    #all
      nrow(tblAnalysis) #2458
    #subset
      nrow(unique(tblAnalysis[,c('sampleID','analysisDate','frozen','filtered')])) #2458
  #merge to df Data
    a<-merge(Data,tblAnalysis,by=c('sampleID','analysisDate','frozen','filtered'))
    Data<-merge(Data,a[,c('ID','analysisID')],by='ID')

```

**build tblFluorometry**

* Rename fields to match cyanoMon.mdb (Old:New)
    - Parameter:parameter
    - Value:value
    - Units:units
    - Fluorometer:fluorometerType
    - Rep:rep
    
* Add placeholder fields
     - commentFluorometry
     
* Get unique values for c('sampleID','fluorometryDate','fluorometryTime','analystName','frozen','filtered','dilution','sampleTemperatureC','photosFluorometry','commentFluorometry')

* number of unique fluorometryID (8456) should equal number of rows in df Data (8400)
* There are repeated lines-exact duplicates
  - VTDEC repeated lines for Data$ID==678 and Data$ID==710
  - VTDEC repeated lines for Data$ID==680 and Data$ID==712
  - NHDES Order=142 and Order=143 (see temp.csv) have repeated values WHY WHY WHY
  - other values to check in temp.csv too.



```{r tblFluorometry, include=FALSE, echo=FALSE, cache=FALSE}
#rename fields
  names(Data)[which(names(Data)=="Parameter")]<-"parameter"
  names(Data)[which(names(Data)=="Value")]<-"Value"
  names(Data)[which(names(Data)=="Units")]<-"units"
  names(Data)[which(names(Data)=="Fluorometer")]<-"fluorometerType"
  names(Data)[which(names(Data)=="Rep")]<-"rep"
  
#add placeholder fields
  Data$commentFluorometry<-NA

#create tblFluorometry
  tblFluorometry<-unique(Data[,c("analysisID","parameter","Value","units","rep","fluorometerType","commentFluorometry")]);nrow(tblFluorometry) #8456

c("analysisID","parameter","Value","units","rep","fluorometerType","commentFluorometry")%in%Names()

#add fluorometryID
  fluorometryID<-1:nrow(tblFluorometry)
  tblFluorometry<-data.frame(fluorometryID,tblFluorometry)

#add fluorometryID to df Data
  #select "by" fields
    #all
      nrow(tblFluorometry) #8456
    #subset
      nrow(unique(tblFluorometry[,c("analysisID","parameter","Value","units","rep","fluorometerType")])) #2458
  #merge to df Data
    a<-merge(Data,tblFluorometry,by=c("analysisID","parameter","Value","units","rep","fluorometerType","commentFluorometry"))
    Data<-merge(Data,a[,c('ID','fluorometryID')],by='ID')

#which samples have multiple fluorometryID
  x<-as.data.frame(table(Data$fluorometryID))
  y<-x[x$Freq>1,1]
  k<-Data$fluorometryID%in%y
  w<-Data[k,]
  w[order(w$fluorometryID),]

write.table(w[order(w$fluorometryID),], file='temp.csv',row.names=F,sep=',') #write to csv


a<-Data[,c("ID","analysisID","parameter","Value","units","rep","fluorometerType","commentFluorometry")]
nrow(a) 

b<-merge(a,tblFluorometry,by=c("analysisID","parameter","Value","units","rep","fluorometerType","commentFluorometry"),all.x=TRUE)


x<-as.data.frame(table(b$fluorometryID))
y<-x[x$Freq>1,1]

length(y)#38

k<-b$fluorometryID%in%y

l<-b[k,]



l[order(l$fluorometryID),]


```

















