---
title: "Analysis"
author: "Bryan Milstead"
date: "March 4, 2015"
output: html_document
---
<!---
use these command instead of the knit icon if you want the data and work loaded into the R workspace
  library(knitr)
    setwd()
  knit('Analysis20150115.rmd')
  -->
  
  

```{r Pck, include=FALSE, echo=FALSE, cache=FALSE} 
  #########function to install (if needed) and load R packages by list
libs<-c("rgdal","sp","knitr","maptools","rgeos") #list of packages to load

installLoad<-function(pck){ #user defined function
    if(!pck%in%installed.packages()){install.packages(pck)}
    require(pck, character.only = TRUE)
  }
lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

```{r loadData, include=FALSE, echo=FALSE, cache=FALSE} 
#get monitoring data
  Data2014<-read.csv("Data2014.csv")

#get location data

```
  
  
  
  
  
  
  
  
  
  
  
  

* How many lakes, sites and samples

```{r Sums, include=TRUE, echo=TRUE, cache=TRUE} 
#get data
  Data2014<-read.csv("Data2014.csv")
#select samples only  
  Data1<-Data2014[Data2014$SampleMethod=='Grab'|Data2014$SampleMethod=='VanDorn'|Data2014$SampleMethod=='Integrated',]
#how many observations
  nrow(Data1) #4728 observation
#how many for Beagle/Unfrozen/Unfiltered
  Samples<-Data1[Data1$Fluorometer=='Beagle' & Data1$Frozen==FALSE & Data1$Filtered==FALSE,]
  nrow(Samples)  #1035

  load('Locations.rda')
  nrow(Locations) #227 locations
  length(unique(Locations$WBID)) #75 lakes


#how many samples by org by SampleLocation
table(Samples$Organization,Samples$SampleLocation,useNA='ifany')

#how many lakes by org
a<-unique(data.frame(Org=Data2014$Organization,WBID=Data2014$WBID))
  table(a$Org)

#samples by depth
a<-as.data.frame(table(Samples$Depth,useNA='ifany'))
hist(Samples$Depth)
a<-Samples$Depth
unique(a)
a[a=='surface']<-0
b<-as.data.frame(table(a))
b<-b[b$Freq>0,]

#samples by SampleMethod
as.data.frame(table(Data2014$SampleMethod,useNA='ifany'))
as.data.frame(table(Samples$SampleMethod,useNA='ifany'))

#samples by Units
as.data.frame(table(Data2014$Units,useNA='ifany'))
as.data.frame(table(Samples$Units,useNA='ifany'))


```
 
  
  plot(Lakes[Lakes@data$WB_ID%in%unique(Locations$WBID),])
  
  
unlist(strsplit(names(NH)[15],'_')) #get the column name
  

  
 

 
#get map of USA states
  #us states from Jane Copeland
  US50wgs84<- readShapePoly('L:/Public/Milstead_Lakes/GIS/StateBoundaries/States.shp', 
                         proj4string=CRS("+proj=longlat +datum=WGS84"))   #read a large shapefile with the state boundaries
  #northeast
  NE<-US50wgs84[US50wgs84@data$STATE_ABBR%in%c('RI','CT','MA','VT','NH','ME'),]
  
  #northeast albers
    #ESRI USA Contiguous Albers Equal Area Conic (used by MRB1 WBIDLakes as AlbersX and AlbersY)
      AlbersContiguous<-CRS('+proj=aea +x_0=0 +y_0=0 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +units=m +datum=NAD83')
    NEalb<-spTransform(NE,AlbersContiguous)
      plot(NEalb)
 
  
  Lakes1<-Lakes[Lakes@data$WB_ID%in%unique(WBID$WBID),]
  a<-gCentroid(Lakes1,byid=TRUE,)
  
  windows(10.5,7)
  plot(NE)
  plot(Lakes1,add=TRUE,col='blue')
  plot(a,add=T,pch=16,col='red')
  
  plot(NE)
  plot(Lakes1,add=TRUE,col='blue')
  plot(Loc,add=T,pch=16,col='red')

nrow(Loc) #227
nrow(Lakes1)  #75


a<-Data2014[Data2014$Fluorometer=='Beagle'|is.na(Data2014$Fluorometer),]
with(a,table(Organization,Units))
